---
title: "Untitled"
output: html_document
date: "2026-07-19"
---

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(writexl)
```

```{r}
 
 
Atlas <- read.csv("AMR_combined_harmonized_v2.csv")

names(Atlas)

head(Atlas)
```
```{r}
combined <- Atlas %>%
  select(
    1:107,
    131:135,
    144:148
  )

head(
  combined
)

colnames(combined)
```
Step 1 — Country → Continent mapping
Apply a standard geoscheme to convert every Country value to Continent. Spot-check the edge cases by hand (Russia, Turkey, territories) rather than trusting an automated library blindly — this is quick but easy to get silently wrong.
```{r}
table(combined$Continent)
table(combined$Country)
```
Step 2 — Organism-level filter
Apply the ≥500 total tested isolates AND ≥2 years with >500 isolates rule. This gives you your candidate organism list.
```{r}
library(dplyr)

# ---- Step 1: Build the crosswalk lookup table ----
# One row per raw organism name -> analysis_group / analysis_unit / tier

organism_lookup <- tribble(
  ~Organism_clean, ~analysis_group, ~analysis_unit, ~tier,

  # --- Acinetobacter baumannii (primary) ---
  "Acinetobacter baumannii", "Acinetobacter baumannii", "Acinetobacter baumannii", "Primary",

  # --- Acinetobacter (secondary / sensitivity) ---
  "Acinetobacter baumannii-calcoaceticus species complex", "Acinetobacter baumannii", "ABC species complex", "Secondary",
  "Acinetobacter spp.", "Acinetobacter baumannii", "Acinetobacter spp. (unresolved)", "Secondary",

  # --- Pseudomonas aeruginosa (primary) ---
  "Pseudomonas aeruginosa", "Pseudomonas aeruginosa", "Pseudomonas aeruginosa", "Primary",

  # --- Pseudomonas (secondary) ---
  "Pseudomonas spp.", "Pseudomonas aeruginosa", "Pseudomonas spp. (unresolved)", "Secondary",

  # --- Enterobacterales: Escherichia ---
  "Escherichia coli", "Enterobacterales - Escherichia", "Escherichia coli", "Primary",
  "Escherichia spp.", "Enterobacterales - Escherichia", "Escherichia spp.", "Primary",

  # --- Enterobacterales: Klebsiella ---
  "Klebsiella pneumoniae", "Enterobacterales - Klebsiella", "Klebsiella pneumoniae", "Primary",
  "Klebsiella oxytoca", "Enterobacterales - Klebsiella", "Klebsiella oxytoca", "Primary",
  "Klebsiella aerogenes", "Enterobacterales - Klebsiella", "Klebsiella aerogenes", "Primary",
  "Klebsiella variicola", "Enterobacterales - Klebsiella", "Klebsiella variicola", "Primary",
  "Klebsiella spp.", "Enterobacterales - Klebsiella", "Klebsiella spp.", "Primary",
  "Klebsiella ornithinolytica", "Enterobacterales - Klebsiella", "Klebsiella ornithinolytica", "Primary",
  "Klebsiella planticola", "Enterobacterales - Klebsiella", "Klebsiella planticola", "Primary",
  "Klebsiella ozaenae", "Enterobacterales - Klebsiella", "Klebsiella ozaenae", "Primary",
  "Klebsiella rhinoscleromatis", "Enterobacterales - Klebsiella", "Klebsiella rhinoscleromatis", "Excluded",

  # --- Enterobacterales: Enterobacter (species-complex merged into parent species) ---
  "Enterobacter cloacae", "Enterobacterales - Enterobacter", "Enterobacter cloacae", "Primary",
  "Enterobacter cloacae species complex", "Enterobacterales - Enterobacter", "Enterobacter cloacae", "Primary",
  "Enterobacter hormaechei", "Enterobacterales - Enterobacter", "Enterobacter hormaechei", "Primary",
  "Enterobacter spp.", "Enterobacterales - Enterobacter", "Enterobacter spp.", "Primary",
  "Enterobacter kobei", "Enterobacterales - Enterobacter", "Enterobacter kobei", "Primary",
  "Enterobacter asburiae", "Enterobacterales - Enterobacter", "Enterobacter asburiae", "Primary",
  "Enterobacter bugandensis", "Enterobacterales - Enterobacter", "Enterobacter bugandensis", "Primary",
  "Enterobacter ludwigii", "Enterobacterales - Enterobacter", "Enterobacter ludwigii", "Primary",
  "Enterobacter roggenkampii", "Enterobacterales - Enterobacter", "Enterobacter roggenkampii", "Primary",
  "Enterobacter xiangfangensis", "Enterobacterales - Enterobacter", "Enterobacter xiangfangensis", "Primary",
  "Enterobacter cancerogenus", "Enterobacterales - Enterobacter", "Enterobacter cancerogenus", "Excluded",
  "Enterobacter intermedium", "Enterobacterales - Enterobacter", "Enterobacter intermedium", "Excluded",
  "Enterobacter liquifaciens", "Enterobacterales - Enterobacter", "Enterobacter liquifaciens", "Excluded",

  # --- Enterobacterales: Serratia (species-complex merged into parent species) ---
  "Serratia marcescens", "Enterobacterales - Serratia", "Serratia marcescens", "Primary",
  "Serratia spp.", "Enterobacterales - Serratia", "Serratia spp.", "Primary",
  "Serratia liquefaciens", "Enterobacterales - Serratia", "Serratia liquefaciens", "Primary",
  "Serratia liquefaciens complex", "Enterobacterales - Serratia", "Serratia liquefaciens", "Primary",
  "Serratia ureilytica", "Enterobacterales - Serratia", "Serratia ureilytica", "Primary",
  "Serratia odorifera", "Enterobacterales - Serratia", "Serratia odorifera", "Primary",
  "Serratia rubidaea", "Enterobacterales - Serratia", "Serratia rubidaea", "Primary",
  "Serratia ficaria", "Enterobacterales - Serratia", "Serratia ficaria", "Excluded",
  "Serratia fonticola", "Enterobacterales - Serratia", "Serratia fonticola", "Excluded",
  "Serratia grimesii", "Enterobacterales - Serratia", "Serratia grimesii", "Excluded",
  "Serratia nematodiphila", "Enterobacterales - Serratia", "Serratia nematodiphila", "Excluded",
  "Serratia plymuthica", "Enterobacterales - Serratia", "Serratia plymuthica", "Excluded",

  # --- Enterobacterales: Citrobacter (species-complex merged into parent species) ---
  "Citrobacter freundii", "Enterobacterales - Citrobacter", "Citrobacter freundii", "Primary",
  "Citrobacter freundii species complex", "Enterobacterales - Citrobacter", "Citrobacter freundii", "Primary",
  "Citrobacter koseri", "Enterobacterales - Citrobacter", "Citrobacter koseri", "Primary",
  "Citrobacter spp.", "Enterobacterales - Citrobacter", "Citrobacter spp.", "Primary",
  "Citrobacter braakii", "Enterobacterales - Citrobacter", "Citrobacter braakii", "Primary",
  "Citrobacter amalonaticus", "Enterobacterales - Citrobacter", "Citrobacter amalonaticus", "Primary",
  "Citrobacter farmeri", "Enterobacterales - Citrobacter", "Citrobacter farmeri", "Primary",
  "Citrobacter sedlakii", "Enterobacterales - Citrobacter", "Citrobacter sedlakii", "Primary",
  "Citrobacter amalonaticus/farmeri", "Enterobacterales - Citrobacter", "Citrobacter amalonaticus/farmeri", "Excluded",
  "Citrobacter gillenii", "Enterobacterales - Citrobacter", "Citrobacter gillenii", "Excluded",
  "Citrobacter murliniae", "Enterobacterales - Citrobacter", "Citrobacter murliniae", "Excluded",
  "Citrobacter werkmanii", "Enterobacterales - Citrobacter", "Citrobacter werkmanii", "Excluded",
  "Citrobacter youngae", "Enterobacterales - Citrobacter", "Citrobacter youngae", "Excluded",

  # --- Enterobacterales: Proteus ---
  "Proteus mirabilis", "Enterobacterales - Proteus", "Proteus mirabilis", "Primary",
  "Proteus vulgaris", "Enterobacterales - Proteus", "Proteus vulgaris", "Primary",
  "Proteus spp.", "Enterobacterales - Proteus", "Proteus spp.", "Primary",
  "Proteus hauseri", "Enterobacterales - Proteus", "Proteus hauseri", "Primary",
  "Proteus vulgaris group", "Enterobacterales - Proteus", "Proteus vulgaris group", "Primary",
  "Proteus penneri", "Enterobacterales - Proteus", "Proteus penneri", "Primary",

  # --- Enterobacterales: Providencia ---
  "Providencia stuartii", "Enterobacterales - Providencia", "Providencia stuartii", "Primary",
  "Providencia rettgeri", "Enterobacterales - Providencia", "Providencia rettgeri", "Primary",
  "Providencia spp.", "Enterobacterales - Providencia", "Providencia spp.", "Primary",
  "Providencia alcalifaciens", "Enterobacterales - Providencia", "Providencia alcalifaciens", "Primary",
  "Providencia rustigianii", "Enterobacterales - Providencia", "Providencia rustigianii", "Excluded",

  # --- Enterobacterales: Morganella ---
  "Morganella morganii", "Enterobacterales - Morganella", "Morganella morganii", "Primary",
  "Morganella spp.", "Enterobacterales - Morganella", "Morganella spp.", "Excluded",

  # --- Enterobacterales: other genera clearing >=50 individually ---
  "Pantoea agglomerans", "Enterobacterales - Other", "Pantoea agglomerans", "Primary",
  "Cronobacter sakazakii", "Enterobacterales - Other", "Cronobacter sakazakii", "Primary",
  "Pluralibacter gergoviae", "Enterobacterales - Other", "Pluralibacter gergoviae", "Primary",
  "Lelliottia amnigena", "Enterobacterales - Other", "Lelliottia amnigena", "Primary",
  "Hafnia alvei", "Enterobacterales - Other", "Hafnia alvei", "Primary",
  "Raoultella spp.", "Enterobacterales - Other", "Raoultella spp.", "Primary",

  # --- Enterobacterales: everything else <50, excluded ---
  "Pantoea ananatis", "Enterobacterales - Other", "Pantoea ananatis", "Excluded",
  "Pantoea anthophila", "Enterobacterales - Other", "Pantoea anthophila", "Excluded",
  "Pantoea calida", "Enterobacterales - Other", "Pantoea calida", "Excluded",
  "Pantoea dispersa", "Enterobacterales - Other", "Pantoea dispersa", "Excluded",
  "Pantoea eucrina", "Enterobacterales - Other", "Pantoea eucrina", "Excluded",
  "Pantoea piersonii", "Enterobacterales - Other", "Pantoea piersonii", "Excluded",
  "Pantoea septica", "Enterobacterales - Other", "Pantoea septica", "Excluded",
  "Pantoea spp.", "Enterobacterales - Other", "Pantoea spp.", "Excluded",
  "Cronobacter spp.", "Enterobacterales - Other", "Cronobacter spp.", "Excluded",
  "Raoultella terrigena", "Enterobacterales - Other", "Raoultella terrigena", "Excluded",
  "Salmonella enterica subsp. enterica serovar typhi", "Enterobacterales - Other", "Salmonella", "Excluded",
  "Salmonella spp.", "Enterobacterales - Other", "Salmonella", "Excluded",
  "Kluyvera ascorbata", "Enterobacterales - Other", "Kluyvera ascorbata", "Excluded",
  "Kosakonia cowanii", "Enterobacterales - Other", "Kosakonia cowanii", "Excluded",
  "Leclercia adecarboxylata", "Enterobacterales - Other", "Leclercia adecarboxylata", "Excluded",
  "Mixta calida", "Enterobacterales - Other", "Mixta calida", "Excluded",
  "Phytobacter diazotrophicus", "Enterobacterales - Other", "Phytobacter diazotrophicus", "Excluded",
  "Pseudocitrobacter faecalis", "Enterobacterales - Other", "Pseudocitrobacter faecalis", "Excluded",
  "Rahnella aquatilis", "Enterobacterales - Other", "Rahnella aquatilis", "Excluded",
  "Yersinia enterocolitica", "Enterobacterales - Other", "Yersinia enterocolitica", "Excluded",
  "Gram negative rods in the Family Enterobacteriaceae", "Enterobacterales - Other", "Unresolved Enterobacterales", "Excluded",

  # --- Acinetobacter: below-threshold, excluded ---
  "Acinetobacter anitratus", "Acinetobacter baumannii", "Acinetobacter anitratus", "Excluded"
)

# ---- Step 2: Join onto the combined dataset ----

combined <- combined %>%
  left_join(organism_lookup, by = "Organism_clean")

# ---- Step 3: Check for anything in the data that didn't match the lookup ----
# (should only be non-target organisms -- e.g. Staph, Strep, Enterococcus, etc.
#  which were never in scope and are expected to show as NA here)

unmatched <- combined %>%
  filter(is.na(analysis_group)) %>%
  distinct(Organism_clean)

print(unmatched)   # sanity check -- confirm this list is ONLY out-of-scope organisms

# ---- Step 4: Filter to Primary + Secondary, drop Excluded / true non-target NAs ----

combined_filtered <- combined %>%
  filter(tier %in% c("Primary", "Secondary"))

# ---- Step 5: Sanity check counts match the crosswalk table ----

combined_filtered %>%
  count(analysis_group, tier, sort = TRUE)

 
```
```{r}
combined_filtered_ID<-combined_filtered %>%
  mutate(ID = row_number())
combined_filtered <- combined_filtered %>%
  mutate(ID = row_number()) %>%
  select(ID, everything(),-Isolate.Id)
#-Isolate.Id
```




```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)

# ============================================================
# STEP 0: Identify antibiotic MIC columns and their _I pairs
# ============================================================

interp_cols <- grep("_I$", colnames(combined_filtered), value = TRUE)
mic_cols    <- sub("_I$", "", interp_cols)
mic_cols    <- mic_cols[mic_cols %in% colnames(combined_filtered)]  # keep valid pairs only
interp_cols <- paste0(mic_cols, "_I")

# ============================================================
# STEP 1a: Reshape wide -> long (one row per isolate x antibiotic)
# ============================================================

mic_long <- combined_filtered %>%
  select(ID, Source.Database, Year, Continent, analysis_unit,
         analysis_group, tier, all_of(mic_cols)) %>%
  pivot_longer(cols = all_of(mic_cols), names_to = "Antibiotic", values_to = "MIC_raw")

interp_long <- combined_filtered %>%
  select(ID, all_of(interp_cols)) %>%
  pivot_longer(cols = all_of(interp_cols), names_to = "Antibiotic", values_to = "CLSI_label") %>%
  mutate(Antibiotic = sub("_I$", "", Antibiotic))

long_data <- mic_long %>%
  left_join(interp_long, by = c("ID", "Antibiotic")) %>%
  filter(!is.na(MIC_raw), MIC_raw != "")   # drop "not tested" cells

# ============================================================
# STEP 1b: Parse MIC strings into censored log2 intervals
#   <=X / <X  -> (-Inf, log2(X)]         left-censored
#   >=X / >X  -> [log2(X), Inf)          right-censored
#   exact X   -> (log2(X)-1, log2(X)]    doubling-dilution interval
#               (an "exact" reading of 4 really means true MIC in (2,4])
# ============================================================

parse_mic <- function(mic_str) {
  s   <- str_trim(as.character(mic_str))
  op  <- case_when(
    str_starts(s, "<=") ~ "le",
    str_starts(s, ">=") ~ "ge",
    str_starts(s, "<")  ~ "lt",
    str_starts(s, ">")  ~ "gt",
    TRUE ~ "exact"
  )
  val     <- suppressWarnings(as.numeric(str_remove(s, "^(<=|>=|<|>)")))
  log2val <- log2(val)

  lower <- case_when(
    op %in% c("le", "lt") ~ -Inf,
    op %in% c("ge", "gt") ~ log2val,
    op == "exact"         ~ log2val - 1
  )
  upper <- case_when(
    op %in% c("le", "lt") ~ log2val,
    op %in% c("ge", "gt") ~ Inf,
    op == "exact"         ~ log2val
  )
  tibble(log2_lower = lower, log2_upper = upper, censor_type = op)
}

long_data <- long_data %>%
  bind_cols(parse_mic(long_data$MIC_raw)) %>%
  filter(!is.na(log2_lower) | !is.na(log2_upper))   # drop unparseable strings

# ============================================================
# STEP 1c: Censoring-aware 2-component EM (per organism-antibiotic group)
# ============================================================

truncated_moments <- function(mu, sigma, lower, upper) {
  alpha <- (lower - mu) / sigma
  beta  <- (upper - mu) / sigma

  phi_a <- ifelse(is.infinite(alpha), 0, dnorm(alpha))
  phi_b <- ifelse(is.infinite(beta), 0, dnorm(beta))
  Phi_a <- ifelse(alpha == -Inf, 0, ifelse(alpha == Inf, 1, pnorm(alpha)))
  Phi_b <- ifelse(beta == -Inf, 0, ifelse(beta == Inf, 1, pnorm(beta)))

  Z <- pmax(Phi_b - Phi_a, 1e-12)

  term1 <- (phi_a - phi_b) / Z
  Ex    <- mu + sigma * term1

  alpha_term <- ifelse(is.infinite(alpha), 0, alpha * phi_a)
  beta_term  <- ifelse(is.infinite(beta), 0, beta * phi_b)

  Var <- sigma^2 * (1 + (alpha_term - beta_term) / Z - term1^2)
  Var <- pmax(Var, 1e-8)

  list(Ex = Ex, Ex2 = Var + Ex^2, Z = Z)
}

fit_censored_em <- function(lower, upper, max_iter = 200, tol = 1e-6) {
  mid <- ifelse(is.finite(lower) & is.finite(upper), (lower + upper) / 2,
         ifelse(is.finite(lower), lower + 1, upper - 1))
  q <- quantile(mid, probs = c(0.3, 0.7), na.rm = TRUE)
  mu1 <- unname(q[1]); mu2 <- unname(q[2])
  sigma1 <- sigma2 <- max(sd(mid, na.rm = TRUE), 0.5)
  pi1 <- pi2 <- 0.5
  loglik_prev <- -Inf

  for (iter in 1:max_iter) {
    m1 <- truncated_moments(mu1, sigma1, lower, upper)
    m2 <- truncated_moments(mu2, sigma2, lower, upper)

    L1 <- pi1 * m1$Z
    L2 <- pi2 * m2$Z
    denom <- pmax(L1 + L2, 1e-300)
    r2 <- L2 / denom
    r1 <- 1 - r2

    loglik <- sum(log(denom))

    pi1 <- mean(r1); pi2 <- mean(r2)
    mu1 <- sum(r1 * m1$Ex) / sum(r1)
    mu2 <- sum(r2 * m2$Ex) / sum(r2)
    var1 <- sum(r1 * (m1$Ex2 - 2*mu1*m1$Ex + mu1^2)) / sum(r1)
    var2 <- sum(r2 * (m2$Ex2 - 2*mu2*m2$Ex + mu2^2)) / sum(r2)
    sigma1 <- sqrt(max(var1, 1e-4))
    sigma2 <- sqrt(max(var2, 1e-4))

    if (mu2 < mu1) {  # keep component 2 = resistant (higher mean) by convention
      tmp<-mu1; mu1<-mu2; mu2<-tmp
      tmp<-sigma1; sigma1<-sigma2; sigma2<-tmp
      tmp<-pi1; pi1<-pi2; pi2<-tmp
      r2 <- 1 - r2
    }

    if (abs(loglik - loglik_prev) < tol) break
    loglik_prev <- loglik
  }

  list(mu_S = mu1, sigma_S = sigma1, mu_R = mu2, sigma_R = sigma2,
       pi_R = pi2, posterior_R = r2, loglik = loglik, iterations = iter)
}

# ============================================================
# STEP 1d: Fit EM per organism-antibiotic group (pooled across source & continent),
#          only for groups meeting the EM-fit threshold (n >= 50)
# ============================================================

em_input <- long_data %>%
  group_by(analysis_unit, Antibiotic) %>%
  filter(n() >= 50) %>%
  ungroup()

group_keys <- em_input %>% distinct(analysis_unit, Antibiotic)
cat("Fitting EM for", nrow(group_keys), "organism-antibiotic pairs...\n")

em_long <- em_input %>%
  group_by(analysis_unit, Antibiotic) %>%
  group_modify(~ {
    fit <- tryCatch(fit_censored_em(.x$log2_lower, .x$log2_upper), error = function(e) NULL)
    if (is.null(fit)) {
      .x$posterior_R <- NA_real_
    } else {
      .x$posterior_R <- fit$posterior_R
    }
    .x
  }) %>%
  ungroup() %>%
  filter(!is.na(posterior_R))

# ============================================================
# STEP 1e: Threshold posterior, clean CLSI labels, compute agreement + kappa
# ============================================================

cohen_kappa <- function(x, y) {
  tab <- table(x, y)
  if (nrow(tab) < 2 | ncol(tab) < 2) return(NA_real_)
  n <- sum(tab)
  po <- sum(diag(tab)) / n
  pe <- sum(rowSums(tab) * colSums(tab)) / n^2
  (po - pe) / (1 - pe)
}

em_long <- em_long %>%
  mutate(
    EM_call = if_else(posterior_R > 0.5, "Resistant", "Susceptible"),
    CLSI_label_clean = case_when(
      str_detect(CLSI_label, regex("resist", ignore_case = TRUE))  ~ "Resistant",
      str_detect(CLSI_label, regex("suscept", ignore_case = TRUE)) ~ "Susceptible",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(CLSI_label_clean))

# ============================================================
# STEP 1f: OUTPUT TABLE — organism x antibiotic x source
# ============================================================

step1_comparison_table <- em_long %>%
  group_by(analysis_unit, Antibiotic, Source.Database) %>%
  summarise(
    N = n(),
    raw_agreement = mean(EM_call == CLSI_label_clean),
    kappa = cohen_kappa(EM_call, CLSI_label_clean),
    .groups = "drop"
  ) %>%
  arrange(kappa)

print(step1_comparison_table, n = 100)
```
```{r}
step1_diagnostics <- em_long %>%
  group_by(analysis_unit, Antibiotic, Source.Database) %>%
  summarise(
    N = n(),
    CLSI_pct_resistant = mean(CLSI_label_clean == "Resistant"),
    EM_pi_R = mean(posterior_R > 0.5),   # approx mixing weight
    raw_agreement = mean(EM_call == CLSI_label_clean),
    kappa = cohen_kappa(EM_call, CLSI_label_clean),
    .groups = "drop"
  ) %>%
  arrange(kappa)

print(step1_diagnostics, n = 50)

#write.csv(step1_diagnostics,"step1_diagnostics.csv")
```
```{r}
step1_flagged <- step1_diagnostics %>%
  mutate(
    em_degenerate = EM_pi_R < 0.05 | EM_pi_R > 0.95,
    clsi_extreme_prevalence = CLSI_pct_resistant < 0.05 | CLSI_pct_resistant > 0.95,
    flag_reason = case_when(
      em_degenerate & clsi_extreme_prevalence ~ "Unimodal population -- 2-cluster GMM not meaningful",
      !em_degenerate & clsi_extreme_prevalence ~ "Extreme prevalence -- kappa unstable, use raw agreement/PABAK instead",
      TRUE ~ "OK for kappa validation"
    )
  )

step1_flagged %>% count(flag_reason)
```

```{r}
# ============================================================
# Apply the three-way flag from the diagnostic check
# ============================================================

step1_flagged <- step1_diagnostics %>%
  mutate(
    em_degenerate = EM_pi_R < 0.05 | EM_pi_R > 0.95,
    clsi_extreme_prevalence = CLSI_pct_resistant < 0.05 | CLSI_pct_resistant > 0.95,
    flag_reason = case_when(
      em_degenerate & clsi_extreme_prevalence ~ "Unimodal population",
      !em_degenerate & clsi_extreme_prevalence ~ "Extreme prevalence (kappa unstable)",
      TRUE ~ "OK for kappa validation"
    )
  )

# ============================================================
# Category 1: "Unimodal population" (n=319)
#   -> 2-component GMM isn't a meaningful model here; kappa is
#      not interpretable, EM should NOT be used to judge this
#      pair's label reliability at all.
#   -> DECISION: for these pairs, skip the EM-vs-label validation
#      entirely and just keep whichever original R/S label the
#      source already provided (ATLAS provider / KEYSTONE hand-
#      collected) -- there's no ambiguity to resolve here since
#      near-unanimous prevalence isn't a labeling reliability
#      question in the first place.
# ============================================================

unimodal_pairs <- step1_flagged %>%
  filter(flag_reason == "Unimodal population") %>%
  select(analysis_unit, Antibiotic, Source.Database) %>%
  mutate(secondary_label_source = "original_CLSI_label",
         validation_note = "Unimodal MIC distribution -- GMM not meaningful, kappa not computed, original label used as-is")

# ============================================================
# Category 2: "Extreme prevalence" (n=227)
#   -> Real bimodal-ish structure MAY exist, but standard kappa
#      is mathematically unstable near extreme base rates (can
#      be negative even with good raw agreement).
#   -> DECISION: use raw_agreement as the decision metric for
#      these instead of kappa. Threshold: raw_agreement >= 0.85
#      as a reasonable bar (adjust if you want something stricter/
#      looser -- there's no ERI-style precedent for this specific
#      number, so treat it as a stated, defensible choice, not
#      a borrowed one).
# ============================================================

extreme_prevalence_pairs <- step1_flagged %>%
  filter(flag_reason == "Extreme prevalence (kappa unstable)") %>%
  mutate(
    secondary_label_source = if_else(raw_agreement >= 0.85,
                                       "original_CLSI_label",
                                       "EM_thresholded_substitute"),
    validation_note = "Extreme prevalence -- evaluated via raw agreement (kappa unreliable at this base rate)"
  ) %>%
  select(analysis_unit, Antibiotic, Source.Database, secondary_label_source, validation_note)

# ============================================================
# Category 3: "OK for kappa validation" (n=707)
#   -> Standard decision rule applies: kappa >= 0.6 keeps the
#      original label; below that, substitute EM's thresholded
#      call and flag for manual CLSI re-verification if time
#      allows.
# ============================================================

ok_pairs <- step1_flagged %>%
  filter(flag_reason == "OK for kappa validation") %>%
  mutate(
    secondary_label_source = if_else(kappa >= 0.6,
                                       "original_CLSI_label",
                                       "EM_thresholded_substitute"),
    validation_note = if_else(kappa >= 0.6,
                                "Kappa >= 0.6 -- label accepted",
                                "Kappa < 0.6 -- flagged for manual CLSI re-verification; EM substitute used in the meantime")
  ) %>%
  select(analysis_unit, Antibiotic, Source.Database, secondary_label_source, validation_note)

# ============================================================
# Combine into one master decision table -- this is what
# Step 8 (aggregation) will join against to know which label
# to use for the binary/secondary track, per organism-
# antibiotic-source group.
# ============================================================

label_decision_table <- bind_rows(unimodal_pairs, extreme_prevalence_pairs, ok_pairs)

# Quick summary of how the decision landed across categories
label_decision_table %>% count(secondary_label_source)

# Save it -- you'll join this onto em_long/long_data at the
# aggregation step to pick the right label per group
#write.csv(label_decision_table, "label_decision_table.csv", row.names = FALSE)
```



Step 3 — Frequency table
Run organism counts unfiltered (no category constraint yet), then overlay your three target categories (Enterobacteriaceae, A. baumannii, P. aeruginosa) as a lens. This tells you: which of the three categories actually clear Step 2, and what your provisional top-N organism list looks like. This is the current hard blocker — nothing else below can start until this exists.
```{r}
# ============================================================
# Exclude the 319 "Unimodal population" pairs from the PRIMARY
# R-MIC pipeline entirely. They stay excluded from AAPC/joinpoint/
# hierarchical model -- not from the raw dataset, just from the
# primary resistance-acceleration analysis.
# ============================================================

# Pairs to exclude (organism-antibiotic-source level)
unimodal_exclusions <- label_decision_table %>%
  filter(validation_note == "Unimodal MIC distribution -- GMM not meaningful, kappa not computed, original label used as-is") %>%
  select(analysis_unit, Antibiotic, Source.Database) %>%
  distinct()

# Apply exclusion to the aggregation input (em_long) before it
# feeds Step 8 (aggregation) / Step 9 (primary hierarchical model)
em_long_primary <- em_long %>%
  anti_join(unimodal_exclusions, by = c("analysis_unit", "Antibiotic", "Source.Database"))

# Quick check on what's left
cat("Organism-antibiotic-source groups before exclusion:",
    n_distinct(em_long %>% select(analysis_unit, Antibiotic, Source.Database)), "\n")
cat("Organism-antibiotic-source groups after exclusion:",
    n_distinct(em_long_primary %>% select(analysis_unit, Antibiotic, Source.Database)), "\n")

# Keep a record of what was excluded and why -- this is your
# methods/limitations documentation, not just a working file
#write.csv(unimodal_exclusions, "excluded_unimodal_pairs.csv", row.names = FALSE)
```

```{r}
# ============================================================
# Confirm the exclusion applied correctly
# ============================================================

# Count of organism-antibiotic-source groups BEFORE exclusion
n_before <- em_long %>%
  distinct(analysis_unit, Antibiotic, Source.Database) %>%
  nrow()

# Count of organism-antibiotic-source groups AFTER exclusion
n_after <- em_long_primary %>%
  distinct(analysis_unit, Antibiotic, Source.Database) %>%
  nrow()

# Count of groups flagged for exclusion
n_excluded <- unimodal_exclusions %>%
  distinct(analysis_unit, Antibiotic, Source.Database) %>%
  nrow()

cat("Groups before exclusion: ", n_before, "\n")
cat("Groups flagged to exclude:", n_excluded, "\n")
cat("Groups after exclusion:  ", n_after, "\n")
cat("Expected after (before - excluded):", n_before - n_excluded, "\n")
cat("Match?", (n_before - n_excluded) == n_after, "\n")

# ============================================================
# Double-check: none of the excluded groups still appear
# in em_long_primary (should return 0 rows)
# ============================================================

leftover_check <- em_long_primary %>%
  distinct(analysis_unit, Antibiotic, Source.Database) %>%
  inner_join(unimodal_exclusions, by = c("analysis_unit", "Antibiotic", "Source.Database"))

cat("Excluded groups still present in em_long_primary (should be 0):", nrow(leftover_check), "\n")

# ============================================================
# Isolate-level check: how many rows (isolates) were dropped,
# not just groups -- useful for the methods writeup
# ============================================================

isolates_before <- nrow(em_long)
isolates_after  <- nrow(em_long_primary)

cat("\nIsolates before exclusion:", isolates_before, "\n")
cat("Isolates after exclusion: ", isolates_after, "\n")
cat("Isolates removed:         ", isolates_before - isolates_after,
    sprintf("(%.1f%% of total)", 100 * (isolates_before - isolates_after) / isolates_before), "\n")
```
```{r}
# ============================================================
# STEP 8 — Aggregate to organism x antibiotic x continent x year cells
#
# INSERT: right after the unimodal-exclusion step (the block that
# produced em_long_primary via anti_join with unimodal_exclusions).
# Do NOT run the label-join code from the last few messages --
# that entire branch is being skipped per the simplification.
#
# Uses em_long_primary as-is: posterior_R (primary signal) and
# CLSI_label_clean (secondary signal, original label, no substitution).
# ============================================================

library(dplyr)

CELL_THRESHOLD <- 50   # decided: organism x antibiotic x continent x year >= 50

# ---- Add a point-estimate log2 MIC per isolate, for the primary
#      "R-cluster mean MIC" signal (posterior-weighted average) ----
# Uses the midpoint of the censored interval; for one-sided censored
# values (>X or <=X with no finite far bound) falls back to the
# finite bound +/- 1 as a conservative point estimate.

em_long_primary <- em_long_primary %>%
  mutate(
    MIC_point = case_when(
      is.finite(log2_lower) & is.finite(log2_upper) ~ (log2_lower + log2_upper) / 2,
      is.finite(log2_lower) & !is.finite(log2_upper) ~ log2_lower + 1,   # right-censored (>X)
      !is.finite(log2_lower) & is.finite(log2_upper) ~ log2_upper - 1,   # left-censored (<=X)
      TRUE ~ NA_real_
    )
  )

# ---- Aggregate: one row per organism x antibiotic x continent x year ----

cell_aggregates <- em_long_primary %>%
  group_by(analysis_unit, Antibiotic, Continent, Year) %>%
  summarise(
    N = n(),

    # PRIMARY signal
    R_proportion_weighted = mean(posterior_R),
    R_mean_MIC_weighted   = sum(posterior_R * MIC_point) / sum(posterior_R),

    # SECONDARY signal (original label, no substitution)
    N_labeled          = sum(!is.na(CLSI_label_clean)),
    binary_R_proportion = mean(CLSI_label_clean == "Resistant", na.rm = TRUE),

    .groups = "drop"
  ) %>%
  filter(N >= CELL_THRESHOLD)

# ---- Quick check on what survived ----
cat("Cells before threshold:",
    em_long_primary %>% distinct(analysis_unit, Antibiotic, Continent, Year) %>% nrow(), "\n")
cat("Cells after >=50 threshold:", nrow(cell_aggregates), "\n")

# Save -- this is the direct input to Step 9 (hierarchical model)
# and Step 10 (joinpoint sensitivity)
#write.csv(cell_aggregates, "cell_aggregates.csv", row.names = FALSE)
```
```{r}
 
em_long_primary<-em_long_primary %>%
  left_join(
    combined_filtered_ID %>% select(ID, Country),
    by = "ID"
  )
```

```{r}
library(dplyr)
library(lme4)

CHECKPOINT_DIR <- "rigi_checkpoints"
dir.create(CHECKPOINT_DIR, showWarnings = FALSE)

# ============================================================
# STEP 8b — Country-level aggregation (feeds the mixed model)
#   No hard N threshold here -- the whole point of the
#   hierarchical model is to let sparse country-years borrow
#   strength via partial pooling, rather than being excluded.
# ============================================================

country_year_aggregates <- em_long_primary %>%
  group_by(analysis_unit, Antibiotic, Continent, Country, Year) %>%
  summarise(
    N = n(),
    R_successes = round(sum(posterior_R)),   # posterior-weighted "success" count
    .groups = "drop"
  ) %>%
  mutate(S_failures = N - R_successes)

saveRDS(country_year_aggregates, file.path(CHECKPOINT_DIR, "country_year_aggregates.rds"))
cat("Step 8b done. Rows:", nrow(country_year_aggregates), "\n")

# ============================================================
# STEP 9 — Primary hierarchical logistic model
#   Fit PER organism-antibiotic pair (same modeling unit as EM).
#   Checkpointed: each pair's fit is saved individually, and
#   already-completed pairs are SKIPPED on rerun -- safe to
#   stop and restart this loop at any point.
# ============================================================

pairs_to_fit <- country_year_aggregates %>%
  distinct(analysis_unit, Antibiotic) %>%
  # only bother fitting pairs with reasonable total volume
  inner_join(
    country_year_aggregates %>%
      group_by(analysis_unit, Antibiotic) %>%
      summarise(total_N = sum(N), n_years = n_distinct(Year), .groups = "drop") %>%
      filter(total_N >= 100, n_years >= 3),
    by = c("analysis_unit", "Antibiotic")
  )

cat("Pairs queued for Step 9 fitting:", nrow(pairs_to_fit), "\n")

# Progress log -- lets you see what's done without re-scanning the folder
log_file <- file.path(CHECKPOINT_DIR, "step9_progress_log.csv")
if (!file.exists(log_file)) {
  write.csv(data.frame(analysis_unit=character(), Antibiotic=character(),
                        status=character(), model_type=character(),
                        timestamp=character()),
            log_file, row.names = FALSE)
}

fit_one_pair <- function(unit, antibiotic) {
  safe_name <- gsub("[^A-Za-z0-9]", "_", paste(unit, antibiotic, sep = "__"))
  ckpt_path <- file.path(CHECKPOINT_DIR, paste0(safe_name, ".rds"))

  # SKIP if already done -- this is what makes the loop resumable
  if (file.exists(ckpt_path)) return(invisible(NULL))

  d <- country_year_aggregates %>%
    filter(analysis_unit == unit, Antibiotic == antibiotic)

  # Standardize year to help convergence
  d$year_c <- d$Year - median(d$Year)

  result <- tryCatch({
    # Try random slope first (country-level trend variation)
    m <- glmer(
      cbind(R_successes, S_failures) ~ year_c * Continent + (1 + year_c | Continent:Country),
      data = d, family = binomial,
      control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
    )
    list(model = m, model_type = "random_slope", converged = TRUE)
  }, warning = function(w) {
    # Fall back to random-intercept-only if slope model doesn't converge cleanly
    m <- tryCatch(
      glmer(cbind(R_successes, S_failures) ~ year_c * Continent + (1 | Continent:Country),
            data = d, family = binomial,
            control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))),
      error = function(e) NULL
    )
    list(model = m, model_type = "random_intercept_fallback", converged = !is.null(m))
  }, error = function(e) {
    list(model = NULL, model_type = "failed", converged = FALSE, error_msg = conditionMessage(e))
  })

  saveRDS(list(analysis_unit = unit, antibiotic = antibiotic,
               data = d, result = result),
          ckpt_path)

  write.table(
    data.frame(analysis_unit = unit, Antibiotic = antibiotic,
               status = if (isTRUE(result$converged)) "OK" else "FAILED",
               model_type = result$model_type,
               timestamp = as.character(Sys.time())),
    log_file, sep = ",", append = TRUE, row.names = FALSE, col.names = FALSE
  )

  invisible(NULL)
}

# ---- Run the loop. Safe to interrupt (Ctrl+C / session end) and
#      re-run this exact block later -- completed pairs are skipped ----
for (i in seq_len(nrow(pairs_to_fit))) {
  fit_one_pair(pairs_to_fit$analysis_unit[i], pairs_to_fit$Antibiotic[i])
  if (i %% 10 == 0) cat("Processed", i, "/", nrow(pairs_to_fit), "pairs\n")
}

cat("Step 9 loop complete (or paused). Check", log_file, "for status.\n")

# ============================================================
# CHECK PROGRESS AT ANY TIME (safe to run mid-loop, in a
# separate console/session, without disturbing the fitting loop):
# ============================================================
# progress <- read.csv("rigi_checkpoints/step9_progress_log.csv")
# table(progress$status)
# table(progress$model_type)

# ============================================================
# RESUME instructions (new session / interrupted run):
#   1. Re-run library(dplyr); library(lme4); CHECKPOINT_DIR line
#   2. Re-run STEP 8b (or: country_year_aggregates <- readRDS(
#        file.path(CHECKPOINT_DIR, "country_year_aggregates.rds")))
#   3. Re-run the "pairs_to_fit" block
#   4. Re-run the for-loop -- it skips anything already saved in
#      rigi_checkpoints/ and only fits what's missing
#   5. Check step9_progress_log.csv for how many succeeded/failed
# ============================================================
```

```{r}
# Quick check: are the failures concentrated in already-known
# low-volume organisms, or scattered unpredictably?

progress <- read.csv("rigi_checkpoints/step9_progress_log.csv")

failed <- progress %>% filter(status == "FAILED")

# How many antibiotics failed per organism -- if an organism has
# ALL its antibiotics failing, that's a volume/coverage problem
# with that organism unit, not a per-drug issue
failed %>%
  count(analysis_unit, sort = TRUE)

# Cross-check against total attempted per organism, to see fail RATE
progress %>%
  group_by(analysis_unit) %>%
  summarise(n_total = n(), n_failed = sum(status == "FAILED")) %>%
  filter(n_failed > 0) %>%
  arrange(desc(n_failed / n_total))
```

```{r}
# ============================================================
# Handle the two systematic-failure organisms + document
# the scattered single failures separately
# ============================================================

# 1. Confirm these two units are simply too thin for the model --
#    check total isolate volume and continent/year spread
country_year_aggregates %>%
  filter(analysis_unit %in% c("Enterobacter xiangfangensis", "Escherichia spp.")) %>%
  group_by(analysis_unit) %>%
  summarise(
    total_isolates = sum(N),
    n_continents = n_distinct(Continent),
    n_years = n_distinct(Year)
  )

# 2. Document all 45 as an explicit exclusion table for the
#    methods/limitations section -- reasons split by pattern
excluded_step9 <- progress %>%
  filter(status == "FAILED") %>%
  mutate(
    exclusion_reason = if_else(
      analysis_unit %in% c("Enterobacter xiangfangensis", "Escherichia spp."),
      "Systematic: entire organism unit failed to converge (likely insufficient continent/year coverage)",
      "Isolated: single organism-antibiotic pair failed to converge, likely sparse-cell instability"
    )
  ) %>%
  select(analysis_unit, Antibiotic, exclusion_reason)

#write.csv(excluded_step9, "step9_excluded_pairs.csv", row.names = FALSE)
```

```{r}
# ============================================================
# EXTRACTION — pull trend estimate + 95% CI per organism-antibiotic-
# continent, from the saved glmer model objects
# ============================================================

library(purrr)

extract_trend_estimates <- function(ckpt_file) {
  obj <- readRDS(ckpt_file)
  m <- obj$result$model

  if (is.null(m) || !isTRUE(obj$result$converged)) return(NULL)

  # Fixed-effect coefficients (log-odds scale)
  fe <- fixef(m)
  vc <- vcov(m)
  continents <- levels(factor(obj$data$Continent))
  base_continent <- continents[1]  # reference level used by the model

  # For each continent, the year slope = main "year_c" effect
  # + that continent's interaction term (0 for the reference level)
  results <- map_dfr(continents, function(cont) {
    if (cont == base_continent) {
      slope <- fe["year_c"]
      se <- sqrt(vc["year_c", "year_c"])
    } else {
      inter_term <- paste0("year_c:Continent", cont)
      if (!(inter_term %in% names(fe))) return(NULL)
      slope <- fe["year_c"] + fe[inter_term]
      se <- sqrt(vc["year_c","year_c"] + vc[inter_term,inter_term] +
                 2*vc["year_c", inter_term])
    }

    tibble(
      analysis_unit = obj$analysis_unit,
      Antibiotic = obj$antibiotic,
      Continent = cont,
      log_odds_slope = unname(slope),
      se = unname(se),
      # Convert to an APC-equivalent: % change in odds per year
      APC_equivalent = 100 * (exp(slope) - 1),
      CI_lower = 100 * (exp(slope - 1.96*se) - 1),
      CI_upper = 100 * (exp(slope + 1.96*se) - 1),
      model_type = obj$result$model_type
    )
  })

  results
}

# Run across all checkpoint files
ckpt_files <- list.files("rigi_checkpoints", pattern = "\\.rds$", full.names = TRUE)
ckpt_files <- ckpt_files[!grepl("country_year_aggregates", ckpt_files)]

cat("Extracting from", length(ckpt_files), "checkpoint files...\n")

rigi_resistance_estimates <- map_dfr(ckpt_files, extract_trend_estimates)

cat("Extracted rows:", nrow(rigi_resistance_estimates), "\n")
#write.csv(rigi_resistance_estimates, "rigi_resistance_estimates.csv", row.names = FALSE)

head(rigi_resistance_estimates, 20)
```

```{r}
# ============================================================
# Re-run extraction, capturing whether the vcov warning fired
# per model, so it's a queryable column instead of console noise
# ============================================================

extract_trend_estimates_flagged <- function(ckpt_file) {
  obj <- readRDS(ckpt_file)
  m <- obj$result$model

  if (is.null(m) || !isTRUE(obj$result$converged)) return(NULL)

  vcov_warning_flag <- FALSE
  vc <- withCallingHandlers(
    vcov(m),
    warning = function(w) {
      if (grepl("not positive definite|falling back", conditionMessage(w))) {
        vcov_warning_flag <<- TRUE
      }
      invokeRestart("muffleWarning")
    }
  )

  fe <- fixef(m)
  continents <- levels(factor(obj$data$Continent))
  base_continent <- continents[1]

  map_dfr(continents, function(cont) {
    if (cont == base_continent) {
      slope <- fe["year_c"]; se <- sqrt(vc["year_c", "year_c"])
    } else {
      inter_term <- paste0("year_c:Continent", cont)
      if (!(inter_term %in% names(fe))) return(NULL)
      slope <- fe["year_c"] + fe[inter_term]
      se <- sqrt(vc["year_c","year_c"] + vc[inter_term,inter_term] +
                 2*vc["year_c", inter_term])
    }
    tibble(
      analysis_unit = obj$analysis_unit, Antibiotic = obj$antibiotic, Continent = cont,
      log_odds_slope = unname(slope), se = unname(se),
      APC_equivalent = 100 * (exp(slope) - 1),
      CI_lower = 100 * (exp(slope - 1.96*se) - 1),
      CI_upper = 100 * (exp(slope + 1.96*se) - 1),
      model_type = obj$result$model_type,
      vcov_unreliable = vcov_warning_flag   # NEW
    )
  })
}

rigi_resistance_estimates <- map_dfr(ckpt_files, extract_trend_estimates_flagged)

cat("Rows with unreliable vcov flag:", sum(rigi_resistance_estimates$vcov_unreliable), "/",
    nrow(rigi_resistance_estimates), "\n")

#write.csv(rigi_resistance_estimates, "rigi_resistance_estimates.csv", row.names = FALSE)
```


```{r}
rigi_resistance_estimates <- rigi_resistance_estimates %>%
  mutate(
    ci_width = CI_upper - CI_lower,
    trend_confidence = case_when(
      CI_lower > 0 ~ "Confident increase",
      CI_upper < 0 ~ "Confident decrease",
      TRUE ~ "Direction uncertain"
    )
  )

rigi_resistance_estimates %>% count(trend_confidence)
```

```{r}
# Is uncertainty concentrated in thin continents/organisms (expected),
# or spread evenly across everything (would suggest a real problem
# with model specification, not just data sparsity)?

rigi_resistance_estimates %>%
  group_by(Continent) %>%
  summarise(n = n(), pct_uncertain = mean(trend_confidence == "Direction uncertain")) %>%
  arrange(desc(pct_uncertain))

rigi_resistance_estimates %>%
  group_by(analysis_unit) %>%
  summarise(n = n(), pct_uncertain = mean(trend_confidence == "Direction uncertain")) %>%
  filter(n >= 5) %>%
  arrange(desc(pct_uncertain)) %>%
  head(20)
```

```{r}
library(dplyr)
library(lme4)

# CHECKPOINT_DIR, country_year_aggregates, and pairs_to_fit should
# already exist from the earlier session -- if starting fresh,
# reload them:
# CHECKPOINT_DIR <- "rigi_checkpoints"
# country_year_aggregates <- readRDS(file.path(CHECKPOINT_DIR, "country_year_aggregates.rds"))
# (pairs_to_fit: re-run the block that built it, from a few messages back)

# ---- Paste the function definition from my last message here ----
fit_one_pair_intercept_only <- function(unit, antibiotic) {
  safe_name <- gsub("[^A-Za-z0-9]", "_", paste(unit, antibiotic, sep = "__"))
  ckpt_path <- file.path(CHECKPOINT_DIR, paste0("v2_", safe_name, ".rds"))
  if (file.exists(ckpt_path)) return(invisible(NULL))

  d <- country_year_aggregates %>% filter(analysis_unit == unit, Antibiotic == antibiotic)
  d$year_c <- d$Year - median(d$Year)

  result <- tryCatch({
    m <- glmer(cbind(R_successes, S_failures) ~ year_c * Continent + (1 | Continent:Country),
               data = d, family = binomial,
               control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
    list(model = m, model_type = "random_intercept", converged = TRUE)
  }, error = function(e) list(model = NULL, model_type = "failed", converged = FALSE))

  saveRDS(list(analysis_unit = unit, antibiotic = antibiotic, data = d, result = result), ckpt_path)
  invisible(NULL)
}

# ---- Run it across every pair (new "v2_" files, old ones untouched) ----
for (i in seq_len(nrow(pairs_to_fit))) {
  fit_one_pair_intercept_only(pairs_to_fit$analysis_unit[i], pairs_to_fit$Antibiotic[i])
  if (i %% 50 == 0) cat("Processed", i, "/", nrow(pairs_to_fit), "\n")
}

# ---- Re-run extraction, pointed at the NEW checkpoint files only ----
ckpt_files_v2 <- list.files(CHECKPOINT_DIR, pattern = "^v2_.*\\.rds$", full.names = TRUE)
rigi_resistance_estimates_v2 <- map_dfr(ckpt_files_v2, extract_trend_estimates_flagged)

rigi_resistance_estimates_v2 <- rigi_resistance_estimates_v2 %>%
  mutate(trend_confidence = case_when(
    CI_lower > 0 ~ "Confident increase",
    CI_upper < 0 ~ "Confident decrease",
    TRUE ~ "Direction uncertain"
  ))

rigi_resistance_estimates_v2 %>% count(trend_confidence)
```


```{r}
# Final resistance-side dataset for Tier 2 reporting
rigi_resistance_final <- rigi_resistance_estimates_v2

rigi_resistance_final %>%
  count(trend_confidence) %>%
  mutate(pct = round(100 * n / sum(n), 1))

#write.csv(rigi_resistance_final, "rigi_resistance_final.csv", row.names = FALSE)
```


```{r}
global_funding_clean <- read.csv("method1_target_by_year_funding.csv") %>%
  filter(Year >= 2014, Year <= 2024) %>%  # drop thin early years + incomplete 2025/2026
  rename(matched_target_category = Category)

global_funding_clean %>%
  group_by(matched_target_category) %>%
  summarise(mean_share = mean(share_pct), median_share = median(share_pct))
```

```{r}
# ============================================================
# Fix: derive target_category directly from analysis_unit,
# using the Enterobacterales genus list from your crosswalk
# ============================================================

enterobacterales_genera <- c("Escherichia", "Klebsiella", "Enterobacter", "Serratia",
                              "Citrobacter", "Proteus", "Providencia", "Morganella",
                              "Pantoea", "Cronobacter", "Pluralibacter", "Lelliottia",
                              "Hafnia", "Raoultella")

organism_level_resistance <- rigi_resistance_final %>%
  filter(trend_confidence != "Direction uncertain") %>%
  mutate(
    target_category = case_when(
      analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
      analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
      grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
      TRUE ~ NA_character_   # catches the secondary/sensitivity units: ABC complex, unresolved spp.
    )
  ) %>%
  filter(!is.na(target_category)) %>%
  group_by(target_category) %>%
  summarise(
    n_confident_estimates = n(),
    n_confident_increase = sum(trend_confidence == "Confident increase"),
    n_confident_decrease = sum(trend_confidence == "Confident decrease"),
    pct_estimates_increasing = 100 * n_confident_increase / n_confident_estimates,
    mean_APC_among_confident = mean(APC_equivalent),
    .groups = "drop"
  )

print(organism_level_resistance)
```


```{r}
rigi_tier1 <- organism_level_resistance %>%
  left_join(
    global_funding_clean %>%
      group_by(matched_target_category) %>%
      summarise(mean_investment_share = mean(share_pct), .groups = "drop"),
    by = c("target_category" = "matched_target_category")
  )

print(rigi_tier1, width = Inf)
```


```{r}
organism_level_resistance_v2 <- rigi_resistance_final %>%
  filter(trend_confidence != "Direction uncertain") %>%
  mutate(target_category = case_when(
    analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
    analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
    grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(target_category)) %>%
  group_by(target_category) %>%
  summarise(
    n_confident = n(),
    pct_increasing = 100 * mean(trend_confidence == "Confident increase"),
    mean_APC_among_increasing = mean(APC_equivalent[trend_confidence == "Confident increase"]),
    mean_APC_among_decreasing = mean(APC_equivalent[trend_confidence == "Confident decrease"]),
    .groups = "drop"
  )

print(organism_level_resistance_v2)
```
```{r}
organism_level_resistance_v2 <- organism_level_resistance_v2 %>%
  mutate(
    mean_APC_among_increasing = if_else(is.nan(mean_APC_among_increasing),
                                          NA_real_, mean_APC_among_increasing),
    severity_note = if_else(pct_increasing == 0,
                              "No confidently-increasing cells detected",
                              paste0(round(mean_APC_among_increasing, 1), "% mean APC among increasing cells"))
  )

print(organism_level_resistance_v2)
```
```{r}
enterobacterales_genera <- c("Escherichia", "Klebsiella", "Enterobacter", "Serratia",
                              "Citrobacter", "Proteus", "Providencia", "Morganella",
                              "Pantoea", "Cronobacter", "Pluralibacter", "Lelliottia",
                              "Hafnia", "Raoultella")

rigi_tier2 <- rigi_resistance_final %>%
  filter(trend_confidence != "Direction uncertain") %>%
  mutate(target_category = case_when(
    analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
    analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
    grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(target_category)) %>%
  group_by(target_category, Continent) %>%
  summarise(
    n_confident = n(),
    pct_increasing = 100 * mean(trend_confidence == "Confident increase"),
    mean_APC_among_increasing = mean(APC_equivalent[trend_confidence == "Confident increase"]),
    mean_APC_among_decreasing = mean(APC_equivalent[trend_confidence == "Confident decrease"]),
    .groups = "drop"
  ) %>%
  arrange(target_category, desc(pct_increasing))

print(rigi_tier2, n = 30)
#write.csv(rigi_tier2, "rigi_tier2_regional.csv", row.names = FALSE)
```
```{r}
# ============================================================
# Check whether North America's 86.4% mean APC (Enterobacteriaceae,
# increasing cells) is a broad pattern or driven by a few outliers
# ============================================================

na_entero_increasing <- rigi_resistance_final %>%
  filter(trend_confidence == "Confident increase", Continent == "North America") %>%
  mutate(target_category = case_when(
    analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
    analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
    grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
    TRUE ~ NA_character_
  )) %>%
  filter(target_category == "Enterobacteriaceae") %>%
  arrange(desc(APC_equivalent))

# How many cells, and what does the distribution actually look like?
cat("N cells:", nrow(na_entero_increasing), "\n")
summary(na_entero_increasing$APC_equivalent)

# Top 10 highest APC values -- are a few extreme values pulling the mean up?
print(na_entero_increasing %>% select(analysis_unit, Antibiotic, APC_equivalent, CI_lower, CI_upper) %>% head(10))

# Median vs mean -- if median is much lower than the 86.4% mean,
# that's the tell of a right-skewed distribution driven by outliers
cat("\nMean APC:", mean(na_entero_increasing$APC_equivalent), "\n")
cat("Median APC:", median(na_entero_increasing$APC_equivalent), "\n")

# Recompute mean EXCLUDING the single most extreme value, as a
# quick robustness check
if (nrow(na_entero_increasing) > 1) {
  trimmed_mean <- na_entero_increasing %>%
    slice(-1) %>%   # drop the single highest value
    summarise(mean_APC_excl_top1 = mean(APC_equivalent))
  print(trimmed_mean)
}
```

```{r}
# Add a CI-width based stability flag on top of the existing
# direction-based confidence classification

rigi_resistance_final <- rigi_resistance_final %>%
  mutate(
    ci_width = CI_upper - CI_lower,
    stable_estimate = trend_confidence != "Direction uncertain" & ci_width < 100  # threshold: adjust as needed
  )

# Re-check Enterobacteriaceae/North America using median instead of
# mean, and also restricted to "stable" estimates only

rigi_resistance_final %>%
  filter(trend_confidence == "Confident increase", Continent == "North America") %>%
  mutate(target_category = case_when(
    analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
    analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
    grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
    TRUE ~ NA_character_
  )) %>%
  filter(target_category == "Enterobacteriaceae") %>%
  summarise(
    n = n(),
    n_stable = sum(stable_estimate),
    mean_all = mean(APC_equivalent),
    median_all = median(APC_equivalent),
    mean_stable_only = mean(APC_equivalent[stable_estimate]),
    median_stable_only = median(APC_equivalent[stable_estimate])
  )
```


```{r}
# ============================================================
# Regenerate Tier 1 (organism-level) and Tier 2 (continent-level)
# using median APC + stability filter, uniformly
# ============================================================

rigi_resistance_stable <- rigi_resistance_final %>%
  mutate(
    ci_width = CI_upper - CI_lower,
    stable_estimate = trend_confidence != "Direction uncertain" & ci_width < 100
  ) %>%
  mutate(target_category = case_when(
    analysis_unit == "Acinetobacter baumannii" ~ "Acinetobacter baumannii",
    analysis_unit == "Pseudomonas aeruginosa" ~ "Pseudomonas aeruginosa",
    grepl(paste(enterobacterales_genera, collapse = "|"), analysis_unit) ~ "Enterobacteriaceae",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(target_category))

# Tier 1, redone
rigi_tier1_v2 <- rigi_resistance_stable %>%
  filter(stable_estimate) %>%
  group_by(target_category) %>%
  summarise(
    n_stable = n(),
    pct_increasing = 100 * mean(trend_confidence == "Confident increase"),
    median_APC_increasing = median(APC_equivalent[trend_confidence == "Confident increase"]),
    median_APC_decreasing = median(APC_equivalent[trend_confidence == "Confident decrease"]),
    .groups = "drop"
  )
print(rigi_tier1_v2)

# Tier 2, redone
rigi_tier2_v2 <- rigi_resistance_stable %>%
  filter(stable_estimate) %>%
  group_by(target_category, Continent) %>%
  summarise(
    n_stable = n(),
    pct_increasing = 100 * mean(trend_confidence == "Confident increase"),
    median_APC_increasing = median(APC_equivalent[trend_confidence == "Confident increase"]),
    median_APC_decreasing = median(APC_equivalent[trend_confidence == "Confident decrease"]),
    .groups = "drop"
  ) %>%
  arrange(target_category, desc(pct_increasing))
print(rigi_tier2_v2, n = 30)
```

```{r}
# ============================================================
# Table 3 — final kappa/agreement summary, by source, restricted
# to pairs where kappa is actually a meaningful metric
# ============================================================

table3_summary <- step1_flagged %>%
  filter(flag_reason == "OK for kappa validation") %>%
  group_by(Source.Database) %>%
  summarise(
    n_pairs = n(),
    median_kappa = median(kappa, na.rm = TRUE),
    mean_kappa = mean(kappa, na.rm = TRUE),
    pct_kappa_above_0.6 = 100 * mean(kappa >= 0.6, na.rm = TRUE),
    median_raw_agreement = median(raw_agreement, na.rm = TRUE)
  )

print(table3_summary)

# Secondary: the "Extreme prevalence" subset, reported via raw
# agreement instead of kappa (per the earlier diagnosis that
# kappa is unstable at extreme base rates)
table3_extreme_prevalence <- step1_flagged %>%
  filter(flag_reason == "Extreme prevalence (kappa unstable)") %>%
  group_by(Source.Database) %>%
  summarise(
    n_pairs = n(),
    median_raw_agreement = median(raw_agreement, na.rm = TRUE)
  )

print(table3_extreme_prevalence)
```
```{r}
# ============================================================
# Export all RIGI tables to a single Excel workbook, one sheet
# per table, with footnotes below each table
# ============================================================

if (!require("openxlsx")) install.packages("openxlsx")
library(openxlsx)

wb <- createWorkbook()

write_table_with_footnote <- function(wb, sheet_name, data, footnote) {
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet_name, data, startRow = 1)
  footnote_row <- nrow(data) + 3
  writeData(wb, sheet_name, footnote, startRow = footnote_row, startCol = 1)
  addStyle(wb, sheet_name, style = createStyle(textDecoration = "italic", fontSize = 9),
           rows = footnote_row, cols = 1)
  setColWidths(wb, sheet_name, cols = 1:ncol(data), widths = "auto")
}

# Table 2 — cell coverage (rebuild from cell_aggregates if not already saved)
table2 <- cell_aggregates %>%
  group_by(Continent) %>%
  summarise(n_cells = n(), n_organism_antibiotic_pairs = n_distinct(paste(analysis_unit, Antibiotic))) %>%
  arrange(desc(n_cells))
write_table_with_footnote(wb, "Table2_cell_coverage", table2,
  "Organism x antibiotic x continent x year cells meeting the N>=50 aggregation threshold.")

# Table 3 — EM validation (both subsets)
write_table_with_footnote(wb, "Table3a_kappa_valid", table3_summary,
  "Restricted to organism-antibiotic-source pairs with genuine bimodal MIC structure (n=707 pairs total, excludes 319 near-unimodal pairs and 227 extreme-prevalence pairs -- see Table3b).")
write_table_with_footnote(wb, "Table3b_extreme_prevalence", table3_extreme_prevalence,
  "Pairs where CLSI-label prevalence is <5% or >95%; kappa is statistically unstable at this base rate, so raw agreement is reported instead.")

# Table 4 — hierarchical model convergence (rebuild from progress log)
table4 <- data.frame(
  Metric = c("Organism-antibiotic pairs attempted", "Total estimates extracted",
             "Confident increase", "Confident decrease", "Direction uncertain"),
  Value = c(nrow(pairs_to_fit), nrow(rigi_resistance_final),
            sum(rigi_resistance_final$trend_confidence == "Confident increase"),
            sum(rigi_resistance_final$trend_confidence == "Confident decrease"),
            sum(rigi_resistance_final$trend_confidence == "Direction uncertain"))
)
write_table_with_footnote(wb, "Table4_model_convergence", table4,
  "Trend estimates from the primary hierarchical logistic model (random intercept, country nested in continent). 'Confident' requires the 95% CI to exclude zero.")

# Table 5 — Tier 1 (final, outlier-robust)
write_table_with_footnote(wb, "Table5_Tier1", rigi_tier1_v2,
  "Restricted to stable estimates (direction-confident AND 95% CI width <100 percentage points). Median (not mean) APC used -- see manuscript Methods for rationale. Investment share is mean documented public/philanthropic R&D funding share, 2014-2024, organism-level only (antibiotic-class granularity not supported by GLOBAL data -- see Limitations).")

# Table 6 — Tier 2 (final, outlier-robust)
write_table_with_footnote(wb, "Table6_Tier2", rigi_tier2_v2,
  "Continent-level breakdown, same stability filter and median-based metric as Table 5. Cells with N<10 (e.g. A. baumannii Africa, n=6) should be interpreted cautiously given small sample size.")

saveWorkbook(wb, "RIGI_all_tables.xlsx", overwrite = TRUE)
cat("Saved: RIGI_all_tables.xlsx\n")
```


```{r}
# Confirm A3 isolates are genuinely present throughout, not just
# in the validation step

combined_filtered %>% count(Source.Database)

em_long_primary %>% count(Source.Database)

cell_aggregates_check <- em_long_primary %>%
  distinct(analysis_unit, Antibiotic, Continent, Year) %>%
  nrow()
cat("Sanity: cell_aggregates should be built from all sources pooled -", cell_aggregates_check, "cells\n")
```

